Efficient static analysis of Marlowe contracts

### 2020-05-20 papers/static_analysis2020/paper.tex
Add measurement section mention to introduction
Rewordings and fixes

\maketitle
\begin{abstract}
SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways that properties can be described. However, the way in which properties are described can affect the computational cost of verifying them considerably. In addition, it is quite hard to predict how different changes will affect the computational cost.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item Removing unnecessary parts from the analysis. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item Avoiding high level abstractions. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item Reducing search space by normalising parameters. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the fundamental structure of the implementation. We consider
two main ideas:
\begin{itemize}
\item Reducing search space by using normalised execution paths relevant
to the property. Instead of using search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item Minimizing the representation of inputs and outputs. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the most heavyweight optimisations described in the previous section.

@@

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when
these actions are carried out is taken by the contract on the basis
of the information that is available at each point in time. Participants,
in turn, do not need to be a single person, but they may be represented
by a \texttt{Role}, which in turn would be controlled by another contract.

The Marlowe language is written as a set of mutually recursive Haskell
data types. It regulates interactions between a finite number of participants,
established before the start of the contract.

The main data type is called \texttt{Contract}, and it represents
the logic and actions that the contract allows or enforces. The outmost
constructs of the contract represent the actions that will be enforced
first, and as those constructs become enforced, the contract will
evolve into one of its continuations (sub-contracts), and the process
will continue until only the construct \texttt{Close} remains.

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue,
depending on a given condition that we call an \texttt{Observation}.
The value of an \texttt{Observation} can depend on previous choices
made by participants, on amounts of money and tokens remaining, or
other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for use later.
The value of a \texttt{Value} can depend on the information available
to the contract at the time. For example, it may store the minimum
slot of the slot interval of the current transaction, or the money
available in a particular account at a given point in time, for use
at a later time in the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three types of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in the contract.
The amount is added to one of the accounts of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}. 
\end{itemize}

@@

As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
contracts are passive: they are executed by transactions being submitted
to the blockchain, with the on chain interpreter checking that the
transaction complies with the contract under execution. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their execution is carried out atomically and deterministically.

Each transaction may include a list of inputs, a set of signatures,
minimum and maximum slot numbers, a set of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid, the transaction should have the same effect
at any slot within that slot range (be deterministic). For example,
if a transaction has a minimum slot number which is before a timeout,
and a maximum slot which after a timeout then the transaction will
fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

The key aspect of the Marlowe semantics is that it checks that a particular
transaction is valid given the current state and contract. Because
transactions are deterministic, there should be no reason why someone
accidentally sends a transaction that is invalid, since it will only
result in a cost to that participant. The type signature of the transaction
validation function is:

@@

This function can be factored into four main functions:
\begin{itemize}
\item \texttt{reduceContractStep} - this function simplifies the topmost
construct that does not require an input to be executed (i.e: anything
but a \texttt{When} that has not expired or a \texttt{Close} when
accounts are empty). It only simplifies the \texttt{When} construct
if it has expired (i.e. the timeout specified in the \texttt{When}
is less than or equal to the minimum slot number). In the case of
the \texttt{Close} contract, it only refunds one of the accounts at
a time. 
\item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
repeatedly until it has no further effect (i.e. it reaches a fixed
point). 
\item \texttt{applyInput} - this function processes one input from the head
of the input list. The topmost construct must be a \texttt{When} that
is expecting that particular input and has not expired. 
\item \texttt{applyAllInputs} - this function processes a list of inputs.
It calls \texttt{applyInput} for each of the inputs in the list, and
calls \texttt{reduceContractUntilQuiescent} before and after every
call to \texttt{applyInput}.
\end{itemize}
The \texttt{State} stores information about the amount of money and
tokens in the contract at a given time, together with the choices
made, \texttt{Let} bindings made, and a lower bound for the current
slot:

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item Account identifiers include an account owner; an account owner is
a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together these properties ensure that no money or tokens
are locked in the contract forever.
\item Classification of money and tokens into accounts separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item Marlowe does not perform deposits or payments of negative amounts
of money or tokens. It treats them as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully, it will still wait for a null deposit to
be made, and it will continue as if everything is correct, this way
the expected execution of the contract is disturbed as little as possible.
\item Marlowe contracts have an upper limit in the number of inputs that
they accept throughout their life. This limit is implied by the maximum
number of nested When constructs in the contract, since only one input
per When can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This prevents DoS attacks, and makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

\section{Measurements}

>>>>

\maketitle
\begin{abstract}
SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways those properties can be described. However, the way in which properties are described can affect the computational cost of verifying them considerably. In addition, it is quite hard to predict how different changes will affect the computational cost.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item \textbf{Reducing search space by normalising parameters}. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing search space by using normalised execution paths relevant
to the property}. Instead of using search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

@@

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when of
these actions are carried out is taken by the contract on the basis
of the information that is available at each point in time. Participants
may correspond to either individual public keys or to tokens (\texttt{Role}s).
In turn, \texttt{Role}s may be controlled by separate independent contracts.

The Marlowe language is written as a set of mutually recursive Haskell
data types. Marlowe contracts regulate interactions between a finite number
of participants, determined before the contract is deployed.

The main data type in Marlowe is called \texttt{Contract}, and it represents
the logic and actions that the contract allows or enforces. The outmost
constructs of the \texttt{Contract} represent the actions that will be enforced
first and, as those constructs become enforced, the \texttt{Contract} will
evolve into one of its continuations (sub-contracts), and the process
will continue until only the construct \texttt{Close} remains.

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue
from between two possibilities and depending on a given Boolean condition
that we call an \texttt{Observation}. The value of an \texttt{Observation}
can depend on previous choices made by participants, on amounts of money
and tokens remaining, or other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
Expressions of type \texttt{Value} in Marlowe are evaluated to integer
values and they depend on information available to the contract at the
time of evaluation. For example, a \texttt{Let} construct can be used to
store the minimum slot of the slot interval of the current transaction,
or the amount of money available in a particular account at a given point
in time, for use at a later stage of the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three types of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in the contract.
The amount is added to one of the accounts of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}.
\end{itemize}

@@

As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
contracts are passive: their code is executed as part of the validation of
transactions that are submitted to the blockchain. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their validation is carried out atomically and
deterministically.

Each transaction may include a list of inputs, a set of signatures,
minimum and maximum slot numbers, a set of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

A key aspect of the Marlowe semantics is that it checks that a particular
transaction is valid given the current state and contract. Because
transactions are deterministic, there should be no reason why someone
accidentally sends a transaction that is invalid for a given \texttt{State}
and \texttt{Contract}, since it will only result in a cost to that participant.

It is possible that, due to a race condition, a participant will send a
transaction that no longer applies to a running \texttt{Contract} and
\texttt{State}, but such a transaction would simply be ignored by the
blockchain.

The type signature of the transaction validation function is:

@@

This function can be factored into four main functions:
\begin{itemize}
\item \texttt{reduceContractStep} - this function simplifies the topmost
construct that does not require an input to be executed (i.e: anything
but a \texttt{When} that has not expired or a \texttt{Close} when
accounts are empty). It only simplifies the \texttt{When} construct
if it has expired (i.e. the timeout specified in the \texttt{When}
is less than or equal to the minimum slot number). In the case of
the \texttt{Close} contract, it only refunds one of the accounts at
a time. 
\item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
repeatedly until it has no further effect (i.e. it reaches a fixed
point). 
\item \texttt{applyInput} - this function processes one single input.
The topmost construct must be a \texttt{When} that
is expecting that particular input and has not expired. 
\item \texttt{applyAllInputs} - this function processes a list of inputs.
It calls \texttt{applyInput} for each of the inputs in the list, and
calls \texttt{reduceContractUntilQuiescent} before and after every
call to \texttt{applyInput}.
\end{itemize}
The \texttt{State} stores information about the amount of money and
tokens in the contract at a given time, together with the choices
made, \texttt{Let} bindings made, and a lower bound for the current
slot:

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Account identifiers include an account owner}; an account owner is
a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{Classification of money and tokens into accounts} separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully, it will still wait for a null deposit to
be made, and it will continue as if everything is correct, this way
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested When constructs in the contract, since only one input
per When can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This prevents DoS attacks, and makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

\section{Measurements \label{sec:Measurements} }
